visualFields 1.x: Tools for Analyzing the Field of Vision in \proglang{R}

### 2025-05-10 jss-article-rnw/article.Rnw
Started creating gallery for different plots.

\author{Cord Huchzermeyer\\INSTITUTION
   \And Iván Marín-Franch\\Computational Optometry}
\Plainauthor{Cord Huchzermeyer, Iván Marín-Franch}

@@

@@

library(plotrix)

@@

At the end of the test, the inverse of the contrast threshold, $\Delta L_{\rm threshold}$ at each location is recorded in a decibel scale (dB). This dB scale is normally defined as attenuation from the maximum stimulus possible $\Delta L_{\rm max}$ as:
\begin{equation}
dB = -10\ \rm log \left (\frac{\Delta L_{\rm thr}}{\Delta L_{\rm max}} \right)\ .
\end{equation}
For the most common devices, the Humprhery Field Analyzer and the Octopus 900, the maximum luminance differential $\Delta L_{\rm max}$ is $10000\ (\rm apostilb) / \pi = 3183.1\ cd\ m^{-2}$.

@@

The relationship between stimulus and perception is analyzed using mathematical model. Commonly, these models result in an estimate of a *threshold*, which denotes either the absolute stimulus strength that can just barely be perceived or the smallest difference that can be perceived as different.

@@

\subsection{Measurements
Although usually not seen in visual field data, the most elementary unit of a static visual field test is a subject's response. It consists of the constant stimulus parameters (size, presentation time, background luminance), the stimulus location, the stimulus intensity and whether the subject pressed the button within a predefined time window to indicate that he saw the stimulus.

Below a certain *threshold* luminance, a stimulus is usually not perceivable, above the threshold, the stimulus is usually perceived. However, close to threshold, a stimulus is not always reliably perceived. By definition, the threshold is the luminance of the stimulus where the subject will see the stimulus with a probability of 50%.

The most straightforward to measure such a threshold would be to present stimuli several time at a number different, fixed stimulus strengths. This method is called the method of *constant stimuli* and results in the *psychometric function*, which is usually modeled with a cumulative distribution function (CDF), for example a Gaussian CDF. However, this is not the most effective way to estimate a threshold, because stimulus strength far from the threshold will have probabilities to be seen that are close to either 0% or 100%. Testing these strengths several times does not yield much information.

@@

\subsection{False negatives and false positives

@@

\subsection{Clinical applications of perimetry}
In clinical practice, visual field testing is used to identify decreases in visual function caused by disease. Such *visual field defects* or *scotomas* can be caused by diseases that affect tissue anywhere along the optic pathway. This ranges from the optic media (cornea / lens / vitreous) which transmit light, to the retina, which detects lights and preprocesses visual information, to the optic nerve, which transmits information, and up to the visual cortex.

@@

The environment variable locini contains the number of the first column that contains the actual visual field date. Each column $l_{x}$ corresponds to one of {k} locations in the visual field test. A concise way to select the columns $l_1$ to $l_{k}$ is to use the *getvfcols* function.

@@

There mustn't be any columns after these columns. Use the *vffilter* function to select specific fields and the *vfisvalid* function to check whether a table fulfills the criteria. The latter will also tell you what the problem is when it doesn't.

@@

@@

The visualField package provides functions for more sophisticated plotting, which will be explained later. *These functions automatically take care that right eyes and left eyes are plotted correctly.*

@@

But this analysis requires the use of a *normative dataset* from which to obtain models describing the age effect on sensitivity at each location. These functions are usually linear and those *age linear models* are used to obtain the *age-corrected mean normal visual field*.

The resulting *deviations* from age-corrected mean normal visual field are also measured in dB and they are interpreted as follows. By convention, in the Humphrey Field Analyzer, a negative deviation value means that the sensitivity at that location is below mean normal. In contrast, on Octopus perimeters, negative is above and positive below mean normal (therefore, values are called *defects*). In this package, those differences are called *total deviation* and their interpretation are as for the Humphrey perimeter. In case you use visualFields with data from a different perimeter it is important to know what is the definition of *deviation* and change the sign if necessary.

In eye diseases like glaucoma, general defects that affect the whole visual field are often less relevant than focal scotomas. And often ocular opacities caused by cataract can mask such focal defects. For this reason, a statistical post-hoc correction to *total deviation* values is further applied to obtain *pattern deviation* values. The correction factor is often called the *general height* and for the 24-2 Humphrey is obtained by selecting the 7th greatest total deviation value. The *pattern deviation* values equal the *total deviation* values minus the *general height*.

@@

\subsection{Total deviation}
Total deviation is calculated from the sensitivities and from age-related normal values. These are saved in the environment variable *nv* (normal values).

@@

<<>>= probability maps}
print(gettdp(td)[, getvfcols()[1:10]])
print(getpdp(pd)[, getvfcols()[1:10]])
@

@@

Note that you have to specify what type of data is contained in the visual field tables. The default is set to *total deviation*, `td`.

@@

And a similar hybrid plotbut with pattern deviation values and probabilities.

@@

<<>>= eval = FALSE, include = TRUE}
vfsfa(example2, file = "test.pdf")
vfsfashiny(example2)
@

@@

<<>>= echo=FALSE}

@@

<<>>=, warning=FALSE}

@@

Statistical methods and a series with a minimum number of tests are necessary in order to demonstrate disease progression in the visual fields. Statistical methods for progression in glaucoma that can be subdivided into *event-based* and *trend-based* methods. Furthermore, methods can be subdivided into those that require datasets which describe the "normal fluctuation" in a given population, those that rely on a mathematical model of visual field progression, and those that do not rely on either.

Event-based methods are not implemented in the visualFields package. Briefly, they look for progression between the most recent field and a baseline. Progression is characterized by a decrease in sensitivity with a magnitude that is unlikely (< 5%) in the population under examination (population based data is necessary) and needs to be confirmed in further tests (usually at least two).

@@

The *glr* function performs linear regression on global parameters. In the following example, the patient loses `r # round(abs(glr(getgl(example3))$sl), 2)` dB each year and this change is statistically significant, with a $p$-value < `r # format(ceiling(1000 * glr(getgl(example3))$pval) / 1000, nsmall = 3)`.

@@

<<>>=
vfplotplr(example3)
@
In this graphical display, yellow means that the $p$-value for this location is < 0.05, orange that it is < 0.02, red < 0.01, and dark red < 0.005.

Criteria such as the one suggested earlier based on clusters of locations significantly progressing are useful clinically but do not control for specificity. An alternative method that ensure that specificity of progression is mantained at 95% or any other desired level is Permutation of Pointwise Linear Regression or PoPLR [@IMF:Ole2012] [@IMF:Marin2021]. For this method, all 52 $p$-values are combined into a single $S$-statistic, originally introduced by Fisher [@IMF:Fisher1925]. In order to avoid reliance on population statistics or parametric statistics. (Fisher's original work showed that such statistic followed a chi-square distribution but only if the 52 significance tests were independent of each other and the data was Gaussian, two assumptions that are far from reality.) Thus, random permutations of the series of visual fields are generated and, for each, the $S$-statistic computed. If enough random permutations are performed, the resulting set of $S$-values represent an empirical distribution for the null hypothesis that there is no progression. A combined $p$-value can then be obtained to test this hypothesis against deterioration of the visual field. The interpretation of the $p$-value is as with any significance test.

The *poplr* function yields a large list with many intermediate computations. The main result is the $S$-value and its $p$-value. Note that there is a left tailed (deterioration), a right tailed (improvement) and a two-tailed (visual field not stable) versions in the current implementation.

@@

<<>>= eval = FALSE, include = TRUE}

@@

\subsection{Using a different perimeter}
Different manufacturers have different conventions in perimetry. For example, clinicians working with Humphrey Field Analyzers (Carl-Zeiss-Meditec) are often used to report loss of function as negative values (e.g., MD: mean *deviation* of $-10$ dB), while those working with Octopus perimeters (Haag-Streit) often report loss as a positive value (e.g., MD: mean *defect* of 10 dB). Furthermore, clinicians working with Octopus perimeters often use different grids, even though the 24-2 and 10-2 grids are also available. Usually, fields cannot be directly compared because other setting ($L_{max}$) are also different and because different algorithms are used.

@@

\subsection{Advanced topics}
\subsection{Setting up the environment}
A given visual field program has certain properties that remain unchanged during the test. In the visualFields packages these are saved in an *environment*. After setting this environment, you can only analyse data that is consistent with this environment.

@@

<<>>= getters}

@@

The first decision to make is the structure of visual field result data. In a data frame (and a loadable csv file) the first 4 columns must have the uniquely identifiable key (`id`, `eye`, `date`, `time`). To compute *deviations* from age-corrected mean normal, the age must also be included. And it is recommended to add the information about the reliability indices of the exam, such as false positive rate (`fpr`), false negative rate (`fnr`), and fixation losses (`fl`), although the last two have recently been found not to be "reliable". To design your structure, looking at the datasets included in the visualFields package may be useful.

@@

<<>>=
# read the Octopus G1 coordinates and create the locmap
g1 <- list(coord = read.csv("./g1.csv"), name = "G1", desc = "The G1 location map used by Octopus perimeters", bs = numeric())
str(g1)
# set the G1 locmap
setlocmap(g1)
# create and set the graphical parameters
setgpar(vfgpar(g1$coord))
# read visual fields and plot the sensitivities
vfplot(vfread("./vfg1.csv"), type = "s")
@

>>>>

\author{Cord Huchzermeyer~\orcidlink{0000-0002-0408-8981}\\Department of Ophthalmology, University Hospital Erlangen
   \And Iván Marín-Franch\\Computational Optometry}
\Plainauthor{Cord Huchzermeyer, Iván Marín-Franch}

@@

<<echo=FALSE>>=

library(ggplot2)
library(plotrix)

@@

@@

At the end of the test, the inverse of the contrast threshold, $\Delta L_{\rm threshold}$ at each location is recorded in a decibel scale (dB). This dB scale is normally defined as attenuation from the maximum stimulus possible $\Delta L_{\rm max}$ as:
\begin{equation}
dB = -10\ \rm log \left (\frac{\Delta L_{\rm thr}}{\Delta L_{\rm max}} \right)\ .
\end{equation}
For the most common devices, the Humphrey Field Analyzer and the Octopus 900, the maximum luminance differential $\Delta L_{\rm max}$ is $10000\ (\rm apostilb) / \pi = 3183.1\ cd\ m^{-2}$.

@@

The relationship between stimulus and perception is analyzed using mathematical model. Commonly, these models result in an estimate of a \textbf{threshold}, which denotes either the absolute stimulus strength that can just barely be perceived or the smallest difference that can be perceived as different.

@@

\subsection{Measurements}
Although usually not seen in visual field data, the most elementary unit of a static visual field test is a subject's response. It consists of the constant stimulus parameters (size, presentation time, background luminance), the stimulus location, the stimulus intensity and whether the subject pressed the button within a predefined time window to indicate that he saw the stimulus.

Below a certain \textbf{threshold} luminance, a stimulus is usually not perceivable, above the threshold, the stimulus is usually perceived. However, close to threshold, a stimulus is not always reliably perceived. By definition, the threshold is the luminance of the stimulus where the subject will see the stimulus with a probability of 50\%.

The most straightforward to measure such a threshold would be to present stimuli several time at a number different, fixed stimulus strengths. This method is called the method of \textbf{constant stimuli} and results in the \textbf{psychometric function}, which is usually modeled with a cumulative distribution function (CDF), for example a Gaussian CDF. However, this is not the most effective way to estimate a threshold, because stimulus strength far from the threshold will have probabilities to be seen that are close to either 0\% or 100\%. Testing these strengths several times does not yield much information.

@@

\subsection{False negatives and false positives}

@@

\subsection{Clinical applications of perimetry}
In clinical practice, visual field testing is used to identify decreases in visual function caused by disease. Such \textbf{visual field defects} or \textbf{scotomas} can be caused by diseases that affect tissue anywhere along the optic pathway. This ranges from the optic media (cornea / lens / vitreous) which transmit light, to the retina, which detects lights and preprocesses visual information, to the optic nerve, which transmits information, and up to the visual cortex.

@@

The environment variable locini contains the number of the first column that contains the actual visual field date. Each column $l_{x}$ corresponds to one of {k} locations in the visual field test. A concise way to select the columns $l_1$ to $l_{k}$ is to use the \textbf{getvfcols} function.

@@

There mustn't be any columns after these columns. Use the \textbf{vffilter} function to select specific fields and the \textbf{vfisvalid} function to check whether a table fulfills the criteria. The latter will also tell you what the problem is when it doesn't.

@@

<<fig=TRUE>>=

@@

The visualField package provides functions for more sophisticated plotting, which will be explained later. \textbf{These functions automatically take care that right eyes and left eyes are plotted correctly.}

@@

But this analysis requires the use of a \textbf{normative dataset} from which to obtain models describing the age effect on sensitivity at each location. These functions are usually linear and those \textbf{age linear models} are used to obtain the \textbf{age-corrected mean normal visual field}.

The resulting \textbf{deviations} from age-corrected mean normal visual field are also measured in dB and they are interpreted as follows. By convention, in the Humphrey Field Analyzer, a negative deviation value means that the sensitivity at that location is below mean normal. In contrast, on Octopus perimeters, negative is above and positive below mean normal (therefore, values are called \textbf{defects}). In this package, those differences are called \textbf{total deviation} and their interpretation are as for the Humphrey perimeter. In case you use visualFields with data from a different perimeter it is important to know what is the definition of \textbf{deviation} and change the sign if necessary.

In eye diseases like glaucoma, general defects that affect the whole visual field are often less relevant than focal scotomas. And often ocular opacities caused by cataract can mask such focal defects. For this reason, a statistical post-hoc correction to \textbf{total deviation} values is further applied to obtain \textbf{pattern deviation} values. The correction factor is often called the \textbf{general height} and for the 24-2 Humphrey is obtained by selecting the 7th greatest total deviation value. The \textbf{pattern deviation} values equal the \textbf{total deviation} values minus the \textbf{general height}.

@@

\subsection{Total deviation}
Total deviation is calculated from the sensitivities and from age-related normal values. These are saved in the environment variable \textbf{nv} (normal values).

@@

<<probability maps>>=
print(gettdp(td)[, getvfcols()[1:10]])
print(getpdp(pd)[, getvfcols()[1:10]])
@

@@

Note that you have to specify what type of data is contained in the visual field tables. The default is set to \textbf{total deviation}, `td`.

<<fig=TRUE>>=

@@

And a similar hybrid plot, but with pattern deviation values and probabilities.

<<fig=TRUE>>=

@@

<<eval = FALSE>>=
vfsfa(example2, file = "test.pdf")
vfsfashiny(example2)
@

@@

@@

<<warning=FALSE, fig=TRUE>>=

@@

Statistical methods and a series with a minimum number of tests are necessary in order to demonstrate disease progression in the visual fields. Statistical methods for progression in glaucoma that can be subdivided into \textbf{event-based} and \textbf{trend-based} methods. Furthermore, methods can be subdivided into those that require datasets which describe the "normal fluctuation" in a given population, those that rely on a mathematical model of visual field progression, and those that do not rely on either.

Event-based methods are not implemented in the visualFields package. Briefly, they look for progression between the most recent field and a baseline. Progression is characterized by a decrease in sensitivity with a magnitude that is unlikely (< 5\%) in the population under examination (population based data is necessary) and needs to be confirmed in further tests (usually at least two).

@@

The \textbf{glr} function performs linear regression on global parameters. In the following example, the patient loses `r round(abs(glr(getgl(example3))$sl), 2)` dB each year and this change is statistically significant, with a $p$-value < `r format(ceiling(1000 * glr(getgl(example3))$pval) / 1000, nsmall = 3)`.

@@

<<fig=TRUE>>=

vfplotplr(example3)

@
In this graphical display, yellow means that the $p$-value for this location is < 0.05, orange that it is < 0.02, red < 0.01, and dark red < 0.005.

Criteria such as the one suggested earlier based on clusters of locations significantly progressing are useful clinically but do not control for specificity. An alternative method that ensure that specificity of progression is mantained at 95\% or any other desired level is Permutation of Pointwise Linear Regression or PoPLR [@IMF:Ole2012] [@IMF:Marin2021]. For this method, all 52 $p$-values are combined into a single $S$-statistic, originally introduced by Fisher [@IMF:Fisher1925]. In order to avoid reliance on population statistics or parametric statistics. (Fisher's original work showed that such statistic followed a chi-square distribution but only if the 52 significance tests were independent of each other and the data was Gaussian, two assumptions that are far from reality.) Thus, random permutations of the series of visual fields are generated and, for each, the $S$-statistic computed. If enough random permutations are performed, the resulting set of $S$-values represent an empirical distribution for the null hypothesis that there is no progression. A combined $p$-value can then be obtained to test this hypothesis against deterioration of the visual field. The interpretation of the $p$-value is as with any significance test.

The \textbf{poplr} function yields a large list with many intermediate computations. The main result is the $S$-value and its $p$-value. Note that there is a left tailed (deterioration), a right tailed (improvement) and a two-tailed (visual field not stable) versions in the current implementation.

@@

<<eval = FALSE>>=

@@

\subsection{Using a different perimeter}
Different manufacturers have different conventions in perimetry. For example, clinicians working with Humphrey Field Analyzers (Carl-Zeiss-Meditec) are often used to report loss of function as negative values (e.g., MD: mean \textbf{deviation} of $-10$ dB), while those working with Octopus perimeters (Haag-Streit) often report loss as a positive value (e.g., MD: mean \textbf{defect} of 10 dB). Furthermore, clinicians working with Octopus perimeters often use different grids, even though the 24-2 and 10-2 grids are also available. Usually, fields cannot be directly compared because other setting ($L_{max}$) are also different and because different algorithms are used.

@@

\subsection{Advanced topics}
\subsection{Setting up the environment}
A given visual field program has certain properties that remain unchanged during the test. In the visualFields packages these are saved in an \textbf{environment}. After setting this environment, you can only analyse data that is consistent with this environment.

@@

<<getters>>=

@@

The first decision to make is the structure of visual field result data. In a data frame (and a loadable csv file) the first 4 columns must have the uniquely identifiable key (`id`, `eye`, `date`, `time`). To compute \textbf{deviations} from age-corrected mean normal, the age must also be included. And it is recommended to add the information about the reliability indices of the exam, such as false positive rate (`fpr`), false negative rate (`fnr`), and fixation losses (`fl`), although the last two have recently been found not to be "reliable". To design your structure, looking at the datasets included in the visualFields package may be useful.

@@

<<eval=FALSE>>=
# read the Octopus G1 coordinates and create the locmap
g1 <- list(coord = read.csv("./g1.csv"), name = "G1", desc = "The G1 location map used by Octopus perimeters", bs = numeric())
str(g1)
# set the G1 locmap
setlocmap(g1)
# create and set the graphical parameters
setgpar(vfgpar(g1$coord))
# read visual fields and plot the sensitivities
vfplot(vfread("./vfg1.csv"), type = "s")
@
